Predicting detection probabilities for rare and undersampled North American landbirds

### 2025-07-04 article/article.tex
testing git-latexdiff

	\begin{enumerate}
		\item Accurate assessments of population status and trends rely on accounting for probability of detection.
		However, probability of detection is often unaccounted for or unavailable on a species level. 
		Although methods to estimate detection probability have been developed and implemented for many North American birds, estimates are most accurate for common, easy to detect species. 
		There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
		\item Here, we take advantage of similarities in detection probabilities among phylogenetically-related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using hierarchical Bayesian multi-species removal and distance models. 
		\item 	
		We showed that a multi-species model improved precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
		However, when no data were available for a species, the multi-species models presented here tended to be overly conservative in predicting detectability. 
		\item 	We recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.
	\end{enumerate}

@@

>>>>

		Accurate assessments of population status and trends rely on accounting for probability of detection.
		Testing git-latexdiff.
		However, probability of detection is often unaccounted for or unavailable on a species level. 
		Although methods to estimate detection probability have been developed and implemented for many North American birds, estimates are most accurate for common, easy to detect species. 
		There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
		Here, we take advantage of similarities in detection probabilities among phylogenetically-related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using hierarchical Bayesian multi-species removal and distance models. 
		We showed that a multi-species model improved precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
		However, when no data were available for a species, the multi-species models presented here tended to be overly conservative in predicting detectability. 
		We recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.

@@

\section{Keywords}
\par hierarchical modelling, distance sampling, removal sampling, QPAD, detectability, effective detection radius, cue rate, phylogenetics, bayesian

\section{Lay Summary}
\par Our ability to detect a species is an important metric in conservation decisions and population estimates. 
Thus, it is crucial to have detectability information for as many species as possible.
By leveraging phylogenetic relationships and trait similarities between related species, we built hierarchical Bayesian models to estimate detection probabilities for seven species of rare or undersampled North American landbirds.
We showed that species with very few data points can have reasonable estimates by borrowing information from more data-rich species.
However, species with no data points tended to have very conservative estimates that did not vary significantly from mean detectability of similar species.
We conclude that this modelling technique is useful to estimate detection probabilities for undersampled birds, but researchers should use caution when trying to predict detection probabilities for species with no data.
